Ten years of school closures and consolidations in Hamilton, Canada and the impact on multimodal accessibility

### 2025-01-19 paper/Manuscript.qmd
R2 submitted.

---
title: "10 years of school closures and consolidations: the impact on school accessibility"
bibliography: 
  - "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
format:
  nature-pdf: 
    journal: "default"
    keep-tex: true
editor: source
execute: 
  freeze: auto
author:
  - name: "Anastasia Soukhov"
    email: "soukhoa@mcmaster.ca"
    affiliations:
      - id: 1
        address: "School of Earth, Environment and Society, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
    attributes:
      corresponding: true
  - name: "Christopher D. Higgins"
    email: "cd.higgins@utoronto.ca"
    affiliations:
      - id: 2
        address: "Department of Human Geography, University of Toronto Scarborough, 1265 Military Trail, Toronto, ON, M1C 1A4, Canada"
  - name: "Antonio Páez"
    email: "paezha@mcmaster.ca"
    affiliations:
      - id: 1
        address: "School of Earth, Environment and Society, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
  - name: "Moataz Mohamed"
    email: "mmohame@mcmaster.ca"
    affiliations:
      - id: 3
        address: "Department of Civil Engineering, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
abstract: | 
  | Reducing motorized travel and increasing active travel are top priorities for many communities. However, these priorities clash against many government's reluctance to invest in public services, including schools. Public school boards are increasingly asked to do more with less, a directive often accomplished by closing and consolidating schools. Alas, this policy tends to reduce the proximity of schools for many students, thus reducing the chances for active travel. In this paper we study the City of Hamilton, a mid-sized city in Ontario, Canada, which experienced numerous elementary schools closures between 2011 and 2021. Our analysis evaluates how school closures and consolidations reshaped accessibility for elementary school students by motorized and active travel. In this task we use spatial availability, a singly-constrained multimodal measure of spatial accessibility. This method's proportional allocation mechanisms allow us to make consistent comparisons in accessibility between time periods. Analysis is conducted at the parcel level, considering on-the-ground-capacity (OTGC) of schools and observed home-to-school trip lengths for motorized and active modes. Our findings reveal that overall spatial availability in the city declined, and while the school consolidation policy pursued by the province may have achieved cost savings in facility operation and maintenance, it imposes ongoing costs on students, families, and communities through reduced opportunities for physical activity and entrenched reliance on motorized travel.

@@

  tm_shape(cen_data_2021_hamiltononly |> filter(!is.na(perc_urb))) +
  tm_polygons(col = "perc_urb", border.alpha=0, 
              alph=0.8,
              palette = c("yellow3","yellow4"),
              #labels = c("[0-50%)", "[25-50%)", "[50-75%)","[75-100%]"),
              breaks = c(0,  0.25, .5, .75,  1),
              legend.is.portrait = FALSE,
              style="cont",
              title = "Proportion urban land-use") +

@@

walk_OD_flow <- tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0) +
  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0)+
  tm_shape(od_trips_perc_walk |> mutate(mode = "non-motorized"), bbox=SAvail_i_DA_extra) +
  tm_polygons(col = "TAZOrig_walk_perc", border.alpha = 0.4, border.col = "grey20", lwd=0.2,
              palette = "Greys",
              breaks = c(0, .25, .50, .75, 1),
              title = "Proportion of mode use",
              textNA = "Unavailable",
              colorNA = "white",
              legend.show=FALSE)+
  #tm_shape(SAvail_i_DA_extra) + tm_borders(alpha = 0.4, col = "cornflowerblue", lwd=0.2)+
  tm_shape(ham_taz) + tm_borders(alpha=0.4, col = "grey40", lwd=0.2) +
  tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
  tm_lines(col = "COMMUNITY_", 
           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
           title.col = "Hamilton communities", lwd = 1, legend.col.show=FALSE) + 
  tm_shape(ham_taz_centroids) + tm_dots(alpha=0.5, col="darkblue", size=0.01) +
  tm_shape(desire_lines_ham |> filter(TAZOrig_walk_trips >= 1) 
           |> mutate(mode = "non-motorized")) +
  tm_lines(palette = "Reds", breaks = c(6, 22, 26, 52, 353),
           title.col = "Number of trips (non-motorized)",
           col = "TAZOrig_walk_trips",
           alpha = 0.9,
           scale = 4,
           lwd = "all",
           legend.col.show = FALSE,
           legend.lwd.show = FALSE) +
  tm_facets(by = c("mode","Year"), nrow=1)+
  tmap_defaults +
  tm_layout(bg.color = "grey99",
            panel.label.bg.color = "white",
            panel.label.height = c(0.8,0,0)) +
  # tm_add_legend(type = "line", 
  #               col = c("grey40","lightblue"),
  #               labels = c("TAZ boundary","Community boundary"))+
  tm_add_legend(type = "symbol",
                col = "darkblue",
                labels = "TAZ centroid")

OD_flow_plot <- tmap_arrange(car_OD_flow_plot, walk_OD_flow, nrow = 2)

tmap_save(OD_flow_plot, paste0(here::here(),"/Figures/Fig4.tiff"), height=4.5, width=8, dpi=600)
```

@@

```{r ploting-empirical-and-theo-curve-TLD-tts, eval=FALSE}
TLD_empirical_wtheo_curve_plot <- ggplot() +
  geom_histogram(data = od_trips_perc_tt, 
                 aes(x = x, 
                     y = ..density..),
                 colour = "grey70",
                 fill = "grey70",
                 boundary = 0,
                 binwidth = 3) +
  geom_vline(data = q_1, mapping = aes(xintercept = x), colour = "grey60") +
  geom_vline(data = q_2, mapping = aes(xintercept = x), colour = "grey60") +
  geom_vline(data = q_3, mapping = aes(xintercept = x), colour = "grey60") +
  geom_line(data = tld_theoretical, aes(x = t, y = f), colour = "blue", linewidth=0.8, alpha=0.6) + 
  xlab("Estimated travel time associated with TTS OD flows (minutes)") +
  ylab("Density") +
  scale_x_continuous(breaks = seq(0,45,10), expand=c(0,0))+
  scale_y_continuous(breaks = seq(0,0.12,0.04), limits = c(0,0.12))+
  facet_grid(~mode~year)+
  geom_text(data = q_1, 
            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=4) +
  geom_text(data = q_2,
            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=4) +
  geom_text(data = q_3,
            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=4) +
  theme_classic() +
  theme(axis.title = element_text(size = 11),
        panel.border = element_rect(colour = "grey80", fill=NA),
        strip.text = element_text(size=11)) 
TLD_empirical_wtheo_curve_plot
ggsave(paste0(here::here(),"/Figures/Fig5.tiff"), height=4.5, width=6, dpi=600)
```

@@

```{r, eval=FALSE}
Vim_201120162021_plot <-  
  tm_shape(SAvail_im_DA |> filter(V_im >0), bbox=SAvail_im_DA) +
  tm_polygons(col = "V_im", border.alpha = 0, border.col = "grey20", lwd=0.2,
              palette = "Purples",
              alpha=0.7,
              style="quantile", n=10,
              #labels = c("[>0-0.2)", "[0.2-15.2)", "[15.2-54.8)", "[54.8-1200)"),
              title = expression(V[i]^m),
              textNULL = "Unavailable",
              colorNULL = "white")+
  tm_facets(by = c("mode","year"), nrow=1)+
  tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
  tm_lines(col = "COMMUNITY_URB", 
  palette = c("grey0","grey20","grey40", "grey60",
                       "grey80","grey100"),
           title.col = "Hamilton communities", lwd = 1.5) + 
  tm_shape(ALL_SCHOOLS_Elem_formapping2) +
   tm_symbols(shape = "combo_status",
             size= "OTGC",
             alpha=0.6,
             scale = 0.4,
             shapes = c(1,
                        17,24,
                        15,22,
                        18, 23),
          title.shape = "School status",
          title.size = "OTGC",
          shapes.labels =  c(
          "No change (btwn 2011-21)", 
          "Expanded (btwn 2011-16)", 
          "Relocated + expanded (btwn 2011-16)",
          "Closed (btwn 2011-16)", 
          "Closed (btwn 2016-21)",
          "Opened (btwn 2011-16)", 
          "Opened (btwn 2016-21)")
  ) +
  tm_facets(by=c("Mode","combo_year"))+
  tm_credits(c("60,220 spatial availability \n(99.4% of the 60,567 city-wide OTGC)",
               "58,101 spatial availability\n(99.7% of the 58,287 city-wide OTGC)",
               "57,509 spatial availability\n(99.7% of the 58,697 city-wide OTGC)",

@@

vvim_201120162021_plot <-  
  tm_shape(SAvail_im_DA  |> filter(v_im_kid_popadj >0), bbox=SAvail_im_DA) +
  tm_polygons(col = "v_im_kid_popadj", border.alpha = 0, border.col = "grey20", lwd=0.2,
              palette = "RdYlGn",
              midpoint = 1.0,
              title = expression(v[i]^m),
              style="quantile", n=10,
              # breaks = c(0, 0.61, 0.88, 1.18, 2.2), #note the highest value is ~10.29, I'm breaking the values early for the purpose of visualisation
              # labels = c("[0-0.6)", "[0.6-0.9)", "[0.9-1.2)", "[1.2-10.3)"),
              textNULL = "Unavailable",
              colorNULL = "white")+
  tm_facets(by = c("mode","year"), nrow=1)+
  tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
  tm_lines(col = "COMMUNITY_URB", 
           palette = c("grey0","grey20","grey40", "grey60",
                       "grey80","grey100"),
           title.col = "Hamilton communities", lwd = 1.5, legend.col.show = FALSE) + 
tm_shape(ALL_SCHOOLS_Elem_formapping2) +
  tm_symbols(shape = "combo_status",
             size= "OTGC",
             alpha=0.6,
             scale = 0.4,
             shapes = c(1,
                        17,24,
                        15,22,
                        18, 23),
          title.shape = "School status",
          title.size = "OTGC",
          shapes.labels =  c(
          "No change (btwn 2011-21)", 
          "Expanded (btwn 2011-16)", 
          "Relocated + expanded (btwn 2011-16)",
          "Closed (btwn 2011-16)", 
          "Closed (btwn 2016-21)",
          "Opened (btwn 2011-16)", 
          "Opened (btwn 2016-21)"),
          legend.size.show = FALSE,
          legend.shape.show = FALSE) +
  tm_facets(by=c("Mode","combo_year"))+
  tm_credits(c("1.09 avg. spatial availability per capita",
               "1.05 avg. spatial availability per capita",
               "0.99 avg. spatial availability per capita",
                "0.85 avg. spatial availability per capita",
               "0.71 avg. spatial availability per capita",
               "0.68 avg. spatial availability per capita"),
          position = c("LEFT", "BOTTOM")) +
  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0) +
  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0)+
  tmap_defaults +
  tm_layout(bg.color = "grey99",
            panel.label.bg.color = "white",
            panel.label.height = 0.8)

@@

```{r, eval=FALSE}
#New figure 7 showing the changes -- first motor only
motor_Vi_CHANGE_plot <-  
  tm_shape(comparision_mapping_motorized_Vi |> filter(!st_is_empty(geometry)),
           bbox=SAvail_im_DA_changes_2011_21) +
  tm_polygons(col = "value", border.alpha = 0, border.col = "grey20", lwd=0.2, alpha=0.6,
              palette = "RdBu", midpoint=0,
              style="quantile", n=10,
              #breaks = c(-20, -4, -0.2, 0, 20),
              midpoint = 0,
              #labels = c("[-208 to -4)", "[-4 to -0.2)", "[-0.2 to 0)","[0 to 331]"),
              title = expression("Change in "*V[i]^{Motorized}),
              colorNA = "white")+
  tm_facets(by = c("year"), nrow=1)+
  tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
  tm_lines(col = "COMMUNITY_URB", 
           palette = c("grey0","grey20","grey40", "grey60","grey80","grey100"),
           title.col = "Hamilton communities", lwd = 1.5
           #legend.col.show = FALSE
           ) + 
  tm_shape(ALL_SCHOOLS_Elem_formapping3)+
  tm_symbols(shape = "combo_status",
             alpha=0.6,
             scale = 0.3,
             shapes = c(17,24,
                        15,22,
                        18, 23),
          title.shape = "Schools that changed",
          shapes.labels =  c(
          "Expanded (btwn 2011-16)",
          "Relocated + expanded (btwn 2011-16)",
          "Closed (btwn 2011-16)",
          "Closed (btwn 2016-21)",
          "Opened (btwn 2011-16)",
          "Opened (btwn 2016-21)")
          #legend.size.show = FALSE,
          #legend.shape.show = FALSE
          ) +
  tm_facets(by = c("combo_change_year"), nrow=1)+
  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0) + 
  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0)+
  tmap_defaults +
  tm_layout( bg.color = "grey99",
             panel.label.bg.color = "white",
             panel.label.height = 0.8,
             legend.outside = TRUE,
             panel.labels = c("2011-2016", "2016-2021", "2011-2021"))

motor_vvi_CHANGE_plot <-  
  tm_shape(comparision_mapping_motorized_vvi |> filter(!st_is_empty(geometry)),
           bbox=SAvail_im_DA_changes_2011_21) +
  tm_polygons(col = "value", border.alpha = 0, border.col = "grey20", lwd=0.2, alpha=0.6,
              palette = "RdBu", midpoint=0,
              style="quantile", n=10,
              #breaks = c(-20, -4, -0.2, 0, 20),
              midpoint = 0,
              #labels = c("[-208 to -4)", "[-4 to -0.2)", "[-0.2 to 0)","[0 to 331]"),
              title = expression("Change in "*v[i]^{Motorized}),
              colorNA = "white")+
  tm_facets(by = c("year"), nrow=1)+
  tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
  tm_lines(col = "COMMUNITY_URB", 
           palette = c("grey0","grey20","grey40", "grey60","grey80","grey100"),
           title.col = "Hamilton communities", lwd = 1.5,
           legend.col.show = FALSE
           ) + 
  tm_shape(ALL_SCHOOLS_Elem_formapping3)+
  tm_symbols(shape = "combo_status",
             alpha=0.6,
             scale = 0.3,
             shapes = c(17,24,
                        15,22,
                        18, 23),
          title.shape = "Schools that changed",
          shapes.labels =  c(
          "Expanded (btwn 2011-16)",
          "Relocated + expanded (btwn 2011-16)",
          "Closed (btwn 2011-16)",
          "Closed (btwn 2016-21)",
          "Opened (btwn 2011-16)",
          "Opened (btwn 2016-21)"),
          legend.shape.show = FALSE
          ) +
  tm_facets(by = c("combo_change_year"), nrow=1)+
  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0) + 
  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0)+
  tmap_defaults +
  tm_layout( bg.color = "grey99",
             panel.label.bg.color = "white",
             panel.label.height = 0.8,
             legend.outside = TRUE,
             panel.show = FALSE)

>>>>

---
title: "Ten years of school closures and consolidations in Hamilton, Canada and the impact on multimodal accessibility"
bibliography: 
  - "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
format:
  nature-pdf: 
    journal: "default"
    keep-tex: true
editor: source
execute: 
  freeze: auto
author:
  - name: "Anastasia Soukhov"
    email: "soukhoa@mcmaster.ca"
    affiliations:
      - id: 1
        address: "School of Earth, Environment and Society, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
    attributes:
      corresponding: true
  - name: "Christopher D. Higgins"
    email: "cd.higgins@utoronto.ca"
    affiliations:
      - id: 2
        address: "Department of Human Geography, University of Toronto Scarborough, 1265 Military Trail, Toronto, ON, M1C 1A4, Canada"
  - name: "Antonio Páez"
    email: "paezha@mcmaster.ca"
    affiliations:
      - id: 1
        address: "School of Earth, Environment and Society, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
  - name: "Moataz Mohamed"
    email: "mmohame@mcmaster.ca"
    affiliations:
      - id: 3
        address: "Department of Civil Engineering, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
abstract: | 
  | Reducing motorized travel and increased active travel are top priorities for many communities. However, these goals are challenged by pressures that make governments reluctant to re/invest in public services, including schools. Public school boards are increasingly asked to do more with less, a directive often accomplished by closing and consolidating schools. Such policies can reduce the proximity of schools for many students, thereby limiting opportunities for active travel. In this paper, we examine the City of Hamilton, a mid-sized city in Ontario, Canada, which saw numerous elementary schools closures between 2011 and 2021. Our analysis evaluates how school closures and consolidations reshaped accessibility for elementary school students, both by motorized and active travel. We use spatial availability, a singly-constrained multimodal measure of spatial accessibility. This method's proportional allocation mechanisms allow us to make consistent comparisons in accessibility between time periods. The analysis is conducted at the parcel level, taking into account the on-the-ground-capacity (OTGC) of schools and observed home-to-school trip lengths for motorized and active modes. Overall, our findings reveal a decline in school spatial availability in the city. While the school closure and consolidation policy pursued by the province may have achieved cost savings in facility operation and maintenance, it imposed ongoing costs on students, families, and communities through reduced opportunities for physical activity and entrenched reliance on motorized travel.

@@

  tm_shape(cen_data_2021_hamiltononly |> filter(!is.na(perc_urb))) +
  tm_polygons(col = "perc_urb", border.alpha=0, 
              alph=0.7,
              palette = c("yellow3","yellow4"),
              #labels = c("[0-50%)", "[25-50%)", "[50-75%)","[75-100%]"),
              breaks = c(0,  0.25, .5, .75,  1),
              legend.is.portrait = FALSE,
              style="cont",
              title = "% Urbanisation") +

@@

walk_OD_flow <- tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0) +
  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0)+
  tm_shape(od_trips_perc_walk |> mutate(mode = "non-motorized"), bbox=SAvail_i_DA_extra) +
  tm_polygons(col = "TAZOrig_walk_perc", border.alpha = 0.4, border.col = "grey20", lwd=0.2,
              palette = "Greys",
              breaks = c(0, .25, .50, .75, 1),
              title = "Proportion of mode use",
              textNA = "Unavailable",
              colorNA = "white",
              legend.show=FALSE)+
  #tm_shape(SAvail_i_DA_extra) + tm_borders(alpha = 0.4, col = "cornflowerblue", lwd=0.2)+
  tm_shape(ham_taz) + tm_borders(alpha=0.4, col = "grey40", lwd=0.2) +
  tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
  tm_lines(col = "COMMUNITY_", 
           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
           title.col = "Hamilton communities", lwd = 1, legend.col.show=FALSE) + 
  tm_shape(ham_taz_centroids) + tm_dots(alpha=0.5, col="darkblue", size=0.01) +
  tm_shape(desire_lines_ham |> filter(TAZOrig_walk_trips >= 1) 
           |> mutate(mode = "non-motorized")) +
  tm_lines(palette = "Reds", breaks = c(6, 22, 26, 52, 353),
           title.col = "Number of trips (non-motorized)",
           col = "TAZOrig_walk_trips",
           alpha = 0.9,
           scale = 4,
           lwd = "all",
           legend.col.show = FALSE,
           legend.lwd.show = FALSE) +
  tm_facets(by = c("mode","Year"), nrow=1)+
  tmap_defaults +
  tm_layout(bg.color = "grey99",
            panel.label.bg.color = "white",
            panel.label.height = c(0.8,0,0))+
  tm_add_legend(type = "symbol",
                col = "darkblue",
                labels = "TAZ centroid",
                size = 0.1)

OD_flow_plot <- tmap_arrange(car_OD_flow_plot,
                             walk_OD_flow, nrow = 2)
OD_flow_plot
tmap_save(OD_flow_plot, paste0(here::here(),"/Figures/Fig4.tiff"), height=4.5, width=8, dpi=600)
```

@@

```{r ploting-empirical-and-theo-curve-TLD-tts, eval=FALSE}
TLD_empirical_wtheo_curve_plot <- ggplot() +
  geom_histogram(data = od_trips_perc_tt, 
                 aes(x = x, 
                     y = ..density..),
                 colour = "grey70",
                 fill = "grey70",
                 boundary = 0,
                 binwidth = 3) +
  geom_vline(data = q_1, mapping = aes(xintercept = x), colour = "grey60") +
  geom_vline(data = q_2, mapping = aes(xintercept = x), colour = "grey60") +
  geom_vline(data = q_3, mapping = aes(xintercept = x), colour = "grey60") +
  geom_line(data = tld_theoretical, aes(x = t, y = f), colour = "blue", linewidth=0.8, alpha=0.6) + 
  xlab("Estimated travel time associated with TTS OD flows (minutes)") +
  ylab("Density") +
  scale_x_continuous(breaks = seq(0,45,10), expand=c(0,0))+
  scale_y_continuous(breaks = seq(0,0.12,0.04), limits = c(0,0.12))+
  facet_grid(~mode~year)+
  geom_text(data = q_1, 
            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=3.5) +
  geom_text(data = q_2,
            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=3.5) +
  geom_text(data = q_3,
            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=3.5) +
  theme_classic() +
  theme(axis.title = element_text(size = 11),
        panel.border = element_rect(colour = "grey80", fill=NA),
        strip.text = element_text(size=11)) 
TLD_empirical_wtheo_curve_plot
ggsave(paste0(here::here(),"/Figures/Fig5.tiff"), height=4.5, width=6, dpi=600)
```

@@

```{r, eval=FALSE}
Vim_201120162021_plot <-  
  tm_shape(SAvail_im_DA |> filter(V_im >0), bbox=SAvail_im_DA) +
  tm_polygons(col = "V_im", border.alpha = 0, border.col = "grey20", lwd=0.2,
              palette = "Purples",
              alpha=0.7,
              style="quantile", n=10,
              #labels = c("[>0-0.2)", "[0.2-15.2)", "[15.2-54.8)", "[54.8-1200)"),
              title = expression(V[i]^m),
              textNULL = "Unavailable",
              colorNULL = "white")+
  tm_facets(by = c("mode","year"), nrow=1)+
  tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
  tm_lines(col = "COMMUNITY_URB", 
           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
           title.col = "Hamilton communities\n(% urbanisation)", lwd = 1) + 
  tm_shape(ALL_SCHOOLS_Elem_formapping2) +
   tm_symbols(shape = "combo_status",
             size= "OTGC",
             alpha=0.8,
             scale = 0.4,
             shapes = c(1,
                        17,24,
                        15,22,
                        18, 23),
          title.shape = "School status",
          title.size = "OTGC",
          shapes.labels =  c(
          "No change (btwn 2011-21)", 
          "Expanded (btwn 2011-16)", 
          "Relocated + expanded (btwn 2011-16)",
          "Closed (btwn 2011-16)", 
          "Closed (btwn 2016-21)",
          "Opened (btwn 2011-16)", 
          "Opened (btwn 2016-21)")
  ) +
  tm_facets(by=c("Mode","combo_year"))+
  tm_credits(c("60,220 spatial availability \n(99.4% of the 60,567 city-wide OTGC)",
               "58,101 spatial availability\n(99.7% of the 58,287 city-wide OTGC)",
               "57,509 spatial availability\n(99.7% of the 58,697 city-wide OTGC)",

@@

vvim_201120162021_plot <-  
  tm_shape(SAvail_im_DA  |> filter(v_im_kid_popadj >0), bbox=SAvail_im_DA) +
  tm_polygons(col = "v_im_kid_popadj", border.alpha = 0, border.col = "grey20", lwd=0.2,
              palette = "RdYlGn",
              midpoint = 1.0,
              title = expression(v[i]^m),
              style="quantile", n=10,
              # breaks = c(0, 0.61, 0.88, 1.18, 2.2), #note the highest value is ~10.29, I'm breaking the values early for the purpose of visualisation
              # labels = c("[0-0.6)", "[0.6-0.9)", "[0.9-1.2)", "[1.2-10.3)"),
              textNULL = "Unavailable",
              colorNULL = "white")+
  tm_facets(by = c("mode","year"), nrow=1)+
  tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
  tm_lines(col = "COMMUNITY_URB", 
           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
           title.col = "Hamilton communities\n(% urbanisation)", lwd = 1, legend.col.show=FALSE) + 
tm_shape(ALL_SCHOOLS_Elem_formapping2) +
  tm_symbols(shape = "combo_status",
             size= "OTGC",
             alpha=0.8,
             scale = 0.4,
             shapes = c(1,
                        17,24,
                        15,22,
                        18, 23),
          title.shape = "School status",
          title.size = "OTGC",
          shapes.labels =  c(
          "No change (btwn 2011-21)", 
          "Expanded (btwn 2011-16)", 
          "Relocated + expanded (btwn 2011-16)",
          "Closed (btwn 2011-16)", 
          "Closed (btwn 2016-21)",
          "Opened (btwn 2011-16)", 
          "Opened (btwn 2016-21)"),
          legend.size.show = FALSE,
          legend.shape.show = FALSE) +
  tm_facets(by=c("Mode","combo_year"))+
  tm_credits(c("1.09 avg. spatial availability per capita",
               "1.05 avg. spatial availability per capita",
               "0.99 avg. spatial availability per capita",
                "0.85 avg. spatial availability per capita",
               "0.71 avg. spatial availability per capita",
               "0.68 avg. spatial availability per capita"),
          position = c("LEFT", "BOTTOM")) +
  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0) +
  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0)+
  tmap_defaults +
  tm_layout(bg.color = "grey99",
            panel.label.bg.color = "white",
            panel.label.height = 0.8)

@@

```{r, eval=FALSE}
#New figure 7 showing the changes -- first motor only
motor_Vi_CHANGE_plot <-  
  tm_shape(comparision_mapping_motorized_Vi |> filter(!st_is_empty(geometry)),
           bbox=SAvail_im_DA_changes_2011_21) +
  tm_polygons(col = "value", border.alpha = 0, border.col = "grey20", lwd=0.2, alpha=0.6,
              palette = "RdBu", midpoint=0,
              style="quantile", n=10,
              #breaks = c(-20, -4, -0.2, 0, 20),
              midpoint = 0,
              #labels = c("[-208 to -4)", "[-4 to -0.2)", "[-0.2 to 0)","[0 to 331]"),
              title = expression("Change in "*V[i]^{Motorized}),
              colorNA = "white")+
  tm_facets(by = c("year"), nrow=1)+
  tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
  tm_lines(col = "COMMUNITY_URB", 
           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
           title.col = "Hamilton communities\n(% urbanisation)", lwd = 1, legend.col.show=FALSE) + 
  tm_shape(ALL_SCHOOLS_Elem_formapping3)+
  tm_symbols(shape = "combo_status",
             alpha=0.7,
             scale = 0.3,
             shapes = c(17,24,
                        15,22,
                        18, 23),
          title.shape = "Schools that changed",
          shapes.labels =  c(
          "Expanded (btwn 2011-16)",
          "Relocated + expanded (btwn 2011-16)",
          "Closed (btwn 2011-16)",
          "Closed (btwn 2016-21)",
          "Opened (btwn 2011-16)",
          "Opened (btwn 2016-21)")
          #legend.size.show = FALSE,
          #legend.shape.show = FALSE
          ) +
  tm_facets(by = c("combo_change_year"), nrow=1)+
  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0) + 
  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0)+
  tmap_defaults +
  tm_layout( bg.color = "grey99",
             panel.label.bg.color = "white",
             panel.label.height = 0.8,
             legend.outside = TRUE,
             panel.labels = c("2011-2016", "2016-2021", "2011-2021"))

motor_vvi_CHANGE_plot <-  
  tm_shape(comparision_mapping_motorized_vvi |> filter(!st_is_empty(geometry)),
           bbox=SAvail_im_DA_changes_2011_21) +
  tm_polygons(col = "value", border.alpha = 0, border.col = "grey20", lwd=0.2, alpha=0.6,
              palette = "RdBu", midpoint=0,
              style="quantile", n=10,
              #breaks = c(-20, -4, -0.2, 0, 20),
              midpoint = 0,
              #labels = c("[-208 to -4)", "[-4 to -0.2)", "[-0.2 to 0)","[0 to 331]"),
              title = expression("Change in "*v[i]^{Motorized}),
              colorNA = "white")+
  tm_facets(by = c("year"), nrow=1)+
  tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
  tm_lines(col = "COMMUNITY_URB", 
           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
           title.col = "Hamilton communities\n(% urbanisation)", lwd = 1, legend.col.show=FALSE) +  
  tm_shape(ALL_SCHOOLS_Elem_formapping3)+
  tm_symbols(shape = "combo_status",
             alpha=0.7,
             scale = 0.3,
             shapes = c(17,24,
                        15,22,
                        18, 23),
          title.shape = "Schools that changed",
          shapes.labels =  c(
          "Expanded (btwn 2011-16)",
          "Relocated + expanded (btwn 2011-16)",
          "Closed (btwn 2011-16)",
          "Closed (btwn 2016-21)",
          "Opened (btwn 2011-16)",
          "Opened (btwn 2016-21)"),
          legend.shape.show = FALSE
          ) +
  tm_facets(by = c("combo_change_year"), nrow=1)+
  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0) + 
  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0)+
  tmap_defaults +
  tm_layout( bg.color = "grey99",
             panel.label.bg.color = "white",
             panel.label.height = 0.8,
             legend.outside = TRUE,
             panel.show = FALSE)

#now repeat, but non-motorized - changes in 2011-2016, 2016 to 2021 and 2011 through 2021
nonmotor_Vi_CHANGE_plot <-  
  tm_shape(comparision_mapping_nonmotorized_Vi |> filter(!st_is_empty(geometry)),
           bbox=SAvail_im_DA_changes_2011_21) +
  tm_polygons(col = "value", border.alpha = 0, border.col = "grey20", lwd=0.2, alpha=0.6,
              palette = "RdBu", midpoint=0,
              style="quantile", n=10,
              #breaks = c(-20, -4, -0.2, 0, 20),
              midpoint = 0,
              #labels = c("[-208 to -4)", "[-4 to -0.2)", "[-0.2 to 0)","[0 to 331]"),
              title = expression("Change in "*V[i]^{Non-motorized}),
              colorNA = "white")+
  tm_facets(by = c("year"), nrow=1)+
  tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING"))+ 
  tm_lines(col = "COMMUNITY_URB", 
           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
           title.col = "Hamilton communities\n(% urbanisation)", lwd = 1, legend.col.show=FALSE) + 
  tm_shape(ALL_SCHOOLS_Elem_formapping3)+
  tm_symbols(shape = "combo_status",
             alpha=0.7,
             scale = 0.3,
             shapes = c(17,24,
                        15,22,
                        18, 23),
          title.shape = "Schools that changed",
          shapes.labels =  c(
          "Expanded (btwn 2011-16)",
          "Relocated + expanded (btwn 2011-16)",
          "Closed (btwn 2011-16)",
          "Closed (btwn 2016-21)",
          "Opened (btwn 2011-16)",
          "Opened (btwn 2016-21)"),
          legend.shape.show = FALSE
          ) +
  tm_facets(by = c("combo_change_year"), nrow=1)+
  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0) + 
  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0)+
  tmap_defaults +
  tm_layout( bg.color = "grey99",
             panel.label.bg.color = "white",
             panel.label.height = 0.8,
             legend.outside = TRUE,
             panel.show = FALSE)
